A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods

### 2025-09-19 2025/RuSCDays/RSD-2025.tex
delete colontitules and add all emails

\begin{document}
%
\title{A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods}
%
\titlerunning{A Parallel Algorithm for Tuning Hyperparameters}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Marina Usova \letter \orcidID{0000-0002-0722-6884} \and
Ilya Lebedev\orcidID{0000-0002-8736-0652}  \and
Anton Shtanyuk\orcidID{0000-0003-1809-7173} \and
Evgeny Kozinov \orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \orcidID{0000-0001-5273-2471} 
}
%
\authorrunning{M. Usova et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{usova@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%(рус.) В статье рассматривается параллельный алгоритм для решения задач глобальной оптимизации с частично определенной целевой функцией вида «черный ящик». Задачи такого вида возникают в процессе выбора оптимальных значений гиперпараметров методов машинного обучения и искусственного интеллекта. Алгоритм использует схему редукции размерности на основе кривых Пеано в сочетании с асинхронной схемой распараллеливания вычислений. Для демонстрации эффективности распараллеливания проведены вычислительные эксперименты в задаче настройки гиперпараметров методов искусственного интеллекта, используемых для предсказания значений временного ряда. 
The paper considers a parallel algorithm for solving global optimization problems with a partially defined objective function of the ``black box'' type. Such problems arise in the process of selecting optimal values of hyperparameters used in machine learning and artificial intelligence methods. The algorithm employs a dimensionality reduction scheme based on Peano curves in combination with an asynchronous scheme of parallel computing. To demonstrate the effectiveness of parallelization, computational experiments were conducted for the task of tuning the hyperparameters of artificial intelligence methods used to predict the values of a time series.

>>>>

\begin{document}
%
\title{A Parallel Algorithm for Solving Global Optimization Problems and Its Application for Tuning Hyperparameters of AI Methods}
%
\titlerunning{ }%A Parallel Algorithm for Tuning Hyperparameters}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Marina Usova \letter \orcidID{0000-0002-0722-6884} \and
Ilya Lebedev \letter \orcidID{0000-0002-8736-0652}  \and
Anton Shtanyuk \letter \orcidID{0000-0003-1809-7173} \and
Evgeny Kozinov \letter \orcidID{0000-0001-6776-0096} \and
Konstantin Barkalov \letter \orcidID{0000-0001-5273-2471} 
}
%
\authorrunning{ }%M. Usova et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia 
\email{usova@itmm.unn.ru},
\email{ilya.lebedev@itmm.unn.ru},
\email{anton.shtanyuk@itmm.unn.ru},
\email{evgeny.kozinov@itmm.unn.ru},
\email{konstantin.barkalov@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%(рус.) В статье рассматривается параллельный алгоритм для решения задач глобальной оптимизации с частично определенной целевой функцией вида «черный ящик». Задачи такого вида возникают в процессе выбора оптимальных значений гиперпараметров методов машинного обучения и искусственного интеллекта. Алгоритм использует схему редукции размерности на основе кривых Пеано в сочетании с асинхронной схемой распараллеливания вычислений. Для демонстрации эффективности распараллеливания проведены вычислительные эксперименты в задаче настройки гиперпараметров методов искусственного интеллекта, используемых для предсказания значений временного ряда. 
The paper considers a parallel algorithm for solving global optimization problems with a partially defined objective function of the ``black box'' type. Such problems arise in the process of selecting optimal values of hyperparameters used in machine learning and artificial intelligence methods. The algorithm employs a dimensionality reduction scheme based on Peano curves in combination with an asynchronous scheme of parallel computing. To demonstrate the effectiveness of parallelization, computational experiments were conducted for the task of tuning the hyperparameters of artificial intelligence methods used to predict the values of a time series.
